The evolution of gene expression in the term placenta of viviparous mammals

### 2016-01-05 manuscript/manuscript.Rnw
finish up bonobo paragraph+fix top50 table

<<top.genes.function,echo=FALSE,results="hide">>=
top.genes.table <- function(species.name,n=50) {
    temp <- 
        combined.fpkm[species==species.name &
                           !(!is.na(human_name) & 
                                 human_name %in% housekeeping.genes.superset[,gene_short_name]) &
                               !grepl("((RNA\\d|MT)-|^RP[SL]\\d+$)",human_name) &
                                   !grepl("(^[Mm][Tt]-|^R[pP][lsLS]\\d+[abcdABCD]?[l]?(?:-.+)?$|rRNA$)",gene_short_name)
                      ][order(-mean_fpkm)][,list(name_or_id,
                                                 human_name,mean_fpkm)]
    ## remove any genes which have a null name_or_id; they are probably from unassembled contigs
    temp <- temp[!(is.na(name_or_id)|name_or_id==""),]
    temp <- temp[!duplicated(name_or_id)][1:n]
    setkey(temp,"human_name")
    temp <- 
        hugo.nomenclature[temp][order(-mean_fpkm),
                                list(name_or_id,`Approved Symbol`,`Approved Name`,mean_fpkm)]
    temp[!is.na(`Approved Name`),
         `Approved Name`:=paste0("\\href{http://www.ensembl.org/Homo_sapiens/Gene/Summary?g=",`Approved Symbol`,"}{",xtable.sanitize.text(`Approved Name`),"}")]
    temp[!is.na(`Approved Symbol`),
         `Approved Symbol`:=paste0("\\href{http://www.ensembl.org/Homo_sapiens/Gene/Summary?g=",`Approved Symbol`,"}{",xtable.sanitize.text(`Approved Symbol`),"}")]
    if (species.name=="homo sapiens") {
        temp[,name_or_id:=paste0("\\href{http://www.ensembl.org/Homo_sapiens/Gene/Summary?g=",name_or_id,"}{",xtable.sanitize.text(name_or_id),"}")]
    } else {
        temp[,name_or_id:=xtable.sanitize.text(name_or_id)]
    }
    setnames(temp,c("Gene symbol","Human Symbol","Human Name","FPKM"))
    if (species.name=="homo sapiens") {
        temp <- temp[,-2,with=FALSE]
    }
    return(temp)
}
genes.table.sanitize <- function(str){
    return(str)
}
xtable.sanitize.text <- function(str){
    result <- str
    result <- gsub("\\\\", "SANITIZE.BACKSLASH", result)
    result <- gsub("$", "\\$", result, fixed = TRUE)
    result <- gsub(">", "$>$", result, fixed = TRUE)
    result <- gsub("<", "$<$", result, fixed = TRUE)
    result <- gsub("|", "$|$", result, fixed = TRUE)
    result <- gsub("{", "\\{", result, fixed = TRUE)
    result <- gsub("}", "\\}", result, fixed = TRUE)
    result <- gsub("%", "\\%", result, fixed = TRUE)
    result <- gsub("&", "\\&", result, fixed = TRUE)
    result <- gsub("_", "\\_", result, fixed = TRUE)
    result <- gsub("#", "\\#", result, fixed = TRUE)
    result <- gsub("^", "\\verb|^|", result, fixed = TRUE)
    result <- gsub("~", "\\~{}", result, fixed = TRUE)
    result <- gsub("SANITIZE.BACKSLASH", "$\\backslash$",
                   result, fixed = TRUE)
    return(result)
}
gene.expression <- function(species.name,gene,complete=FALSE) {
    fpkm <- combined.fpkm[species==species.name &
                              gene_short_name==gene,
                          mean_fpkm]

    if (complete) {
        return(paste0("\\textit{",gene,"}=",
                      formatC(fpkm,digits=0,format="f"),
                      " \\ac{FPKM}"
                      ))
    } else {
        return(fpkm)
    }
}

@@

The top 50 expressed genes as measured by \ac{FPKM} in the term
placenta of \bonobo are shown in \cref{tab:top50_bonobo_placenta}
with a list of all genes in Supplemental
\cref{tab:all_pan_paniscus_placenta}. Expression in the \bonobo
placenta differs somewhat from that of \human, as evidenced by the
makeup of the top 50 genes (\cref{tab:top50_bonobo_placenta}). The highest expression
 in \bonobo is %
\Sexpr{gene.expression("pan paniscus","PRG2",complete=TRUE)}, %
%
 which is involved in insulin
growth factor signaling and angiogenesis during pregnancy \cite{Weyer.ea2011:Placentalregulationofpeptide}. \textit{PRG2} is also expressed at relatively high levels in the
human placenta (\ac{FPKM}=\Sexpr{gene.expression("homo sapiens","PRG2")}; Supplemental \cref{tab:all_homo_sapiens_placenta}), although at
comparatively much lower levels than the \bonobo. A \bonobo gene closely
related to the chorionic somatomammotropins (ENSPTRG00000034560) is also
expressed at high levels %
%
(\ac{FPKM}=\Sexpr{gene.expression("pan paniscus","ENSPTRG00000034560")}); %
%
\cref{tab:top50_bonobo_placenta}). Two other Pan genes included within
this family (ENSPTRG00000030805 which shows similarity to GH2 and
ENSPTRG00000034560) show expression levels of \ac{FPKM}>89. Six
pregnancy-specific beta-1-glycoprotein genes were expressed in the
\bonobo placenta with \ac{FPKM} ranging from 20.885-167.961
(Supplemental \cref{tab:all_pan_paniscus_placenta}). Placental
expression of the genes involved in estrogen and progesterone
production (CYP19A1, HSD3B1 respectively) show lower expression in the
bonobo as compared to humans %
(\Sexpr{gene.expression("pan paniscus","CYP19A1",complete=TRUE)},
\cref{fig:estrogen_evolution_tree};
\Sexpr{gene.expression("pan paniscus","HSD3B1",complete=TRUE)},
\cref{fig:progesterone_evolution_tree} and
\cref{tab:top50_human_placenta}) although the expression is fairly similar to that of 
. In addition, \textit{CGA} is
expressed in the bonobo placenta
(\ac{FPKM}=\Sexpr{gene.expression("pan paniscus","CGA")}), but at much
lower levels than \human as well (\cref{tab:top50_human_placenta}). 
%
Two Ensembl-annotated beta subunits of the chorionic gonadotropin
genes are expressed in \bonobo, one related to CGB and another related
to CGB7; both are not highly expressed (Table 8). Among the other top
50 expressed genes, there is high expression of ribosomal proteins
(e.g., RPL18, RPL23), keratins (e.g., KRT7, KRT19), heat-shock
proteins (e.g., HSPA5, HSPA1A), serpins (SERPINE1, SERPINE2), insulin
growth factor 2 (IGF2) and the hormone adrenomedullin (ADM).

>>>>

<<top.genes.function,echo=FALSE,results="hide">>=
top.genes.table <- function(species.name,n=50) {
    alignment.species <- species.name
    if (alignment.species=="pan paniscus") {
        alignment.species <- "pan troglodytes"
    } else if (alignment.species=="ateles fusciceps") {
        alignment.species <- "homo sapiens"
    } else if (grepl("spalax",alignment.species)) {
        alignment.species <- "mus musculus"
    }
    temp <- 
        combined.fpkm[species==species.name &
                      !(!is.na(human_name) & 
                        human_name %in% housekeeping.genes.superset[,gene_short_name]) &
                      ! (gene_biotype %in% c("Mt_tRNA","Mt_rRNA","rRNA")) &
                      !grepl("((RNA\\d|MT)-|^RP[SL]P?\\d+[ABCD]?$)",human_name) &
                      !grepl("((RNA\\d|MT)-|^RP[SL]P?\\d+[ABCD]?$)",human_alignment_symbol) &
                      !grepl("(^[Mm][Tt]-|^R[pP][lsLS]\\d+[abcdABCD]?[l]?(?:-.+)?$|rRNA$)",
                             gene_short_name)
                      ][order(-mean_fpkm)][,list(name_or_id,
                                                 human_alignment_symbol,
                                                 human_name,mean_fpkm)]
    ## remove any genes which have a null name_or_id; they are probably from unassembled contigs
    temp <- temp[!(is.na(name_or_id)|name_or_id==""),]
    temp <- temp[!duplicated(name_or_id)][1:n]
    temp[,used_alignment:=FALSE]
    temp[is.na(human_name) & !is.na(human_alignment_symbol),
                    used_alignment:=TRUE]
    temp[is.na(human_name),human_name:=human_alignment_symbol]
    setkey(temp,"human_name")
    temp <- 
        hugo.nomenclature[temp][order(-mean_fpkm),
                                list(name_or_id,used_alignment,`Approved Symbol`,`Approved Name`,mean_fpkm)]
    temp[!is.na(name_or_id),
         name_or_id:=paste0("\\href{http://www.ensembl.org/",
                             capfirst(gsub(" ","_",alignment.species)),
                             "/Gene/Summary?g=",name_or_id,"}{",
                             xtable.sanitize.text(name_or_id),"}")]
    temp[!is.na(`Approved Name`),
         `Approved Name`:=paste0("\\href{http://www.ensembl.org/Homo_sapiens/Gene/Summary?g=",`Approved Symbol`,"}{",xtable.sanitize.text(`Approved Name`),"}")]
    temp[!is.na(`Approved Symbol`),
         `Approved Symbol`:=paste0("\\href{http://www.ensembl.org/Homo_sapiens/Gene/Summary?g=",`Approved Symbol`,"}{",xtable.sanitize.text(`Approved Symbol`),"}")]
    temp[used_alignment==TRUE,`Approved Symbol`:=paste0("(",`Approved Symbol`,")")]
    temp[,used_alignment:=NULL]
    setnames(temp,c("Gene symbol","Human Symbol","Human Name","FPKM"))
    if (species.name=="homo sapiens" || species.name=="ateles fusciceps") {
        temp[,`Human Symbol`:=NULL]
    }
    return(temp)
}
genes.table.sanitize <- function(str){
    return(str)
}
xtable.sanitize.text <- function(str){
    result <- str
    result <- gsub("\\\\", "SANITIZE.BACKSLASH", result)
    result <- gsub("$", "\\$", result, fixed = TRUE)
    result <- gsub(">", "$>$", result, fixed = TRUE)
    result <- gsub("<", "$<$", result, fixed = TRUE)
    result <- gsub("|", "$|$", result, fixed = TRUE)
    result <- gsub("{", "\\{", result, fixed = TRUE)
    result <- gsub("}", "\\}", result, fixed = TRUE)
    result <- gsub("%", "\\%", result, fixed = TRUE)
    result <- gsub("&", "\\&", result, fixed = TRUE)
    result <- gsub("_", "\\_", result, fixed = TRUE)
    result <- gsub("#", "\\#", result, fixed = TRUE)
    result <- gsub("^", "\\verb|^|", result, fixed = TRUE)
    result <- gsub("~", "\\~{}", result, fixed = TRUE)
    result <- gsub("SANITIZE.BACKSLASH", "$\\backslash$",
                   result, fixed = TRUE)
    return(result)
}
gene.expression <- function(species.name,gene,complete=FALSE) {
    fpkm <- combined.fpkm[species==species.name &
                              (gene_short_name==gene | 
                               gene_id==gene),
                          mean_fpkm]
    if (complete) {
        if (any(grepl("^ENS.+G\\d+$",gene))) {
            return(paste0(gene,
                          formatC(fpkm,digits=0,format="f"),
                          " \\ac{FPKM}"))
        } else {
            return(paste0("\\textit{",gene,"}=",
                          formatC(fpkm,digits=0,format="f"),
                          " \\ac{FPKM}"
                          ))
        }
    } else {
        return(fpkm)
    }
}

@@

The top 50 expressed genes as measured by \ac{FPKM} in the term
placenta of \bonobo are shown in \cref{tab:top50_bonobo_placenta}
with a list of all genes in Supplemental
\cref{tab:all_pan_paniscus_placenta}. Expression in the \bonobo
placenta differs somewhat from that of \human, as evidenced by the
makeup of the top 50 genes (\cref{tab:top50_bonobo_placenta}). The highest expression
 in \bonobo is %
\Sexpr{gene.expression("pan paniscus","PRG2",complete=TRUE)}, %
%
which is involved in insulin growth factor signaling and angiogenesis
during pregnancy \cite{Weyer.ea2011:Placentalregulationofpeptide}.
\textit{PRG2} is also expressed at relatively high levels (though
comparatively much lower) in the human placenta
(\ac{FPKM}=\Sexpr{gene.expression("homo sapiens","PRG2")};
Supplemental \cref{tab:all_homo_sapiens_placenta}). Four \bonobo genes
related to the chorionic somatomammotropins
are also expressed at relatively high levels (GH1 
\Sexpr{gene.expression("pan paniscus","ENSPTRG00000034560",complete=TRUE)}, 
\Sexpr{gene.expression("pan paniscus","ENSPTRG00000034212",complete=TRUE)}, 
GH2
\Sexpr{gene.expression("pan paniscus","ENSPTRG00000030805",complete=TRUE)}, 
\Sexpr{gene.expression("pan paniscus","ENSPTRG00000034209",complete=TRUE)}, 
%
\cref{tab:all_pan_paniscus_placenta}, \cref{fig:csh1_evolution_tree}).
Placental expression of the genes involved in estrogen and
progesterone production (\textit{CYP19A1}, \textit{HSD3B1}
respectively) show lower expression in \bonobo as compared to \human %
(\Sexpr{gene.expression("pan paniscus","CYP19A1",complete=TRUE)},
\cref{fig:estrogen_evolution_tree}; %
\Sexpr{gene.expression("pan paniscus","HSD3B1",complete=TRUE)},
\cref{fig:progesterone_evolution_tree} and
\cref{tab:top50_human_placenta}) although the expression of
\textit{HSD3B1} is fairly similar to that of \smonkey %
(\Sexpr{gene.expression("ateles fusciceps","HSD3B1",complete=TRUE)}. %
In addition, \textit{CGA} is expressed in the bonobo placenta %
(\ac{FPKM}=\Sexpr{gene.expression("pan paniscus","CGA")}), but at much
lower levels than \human as well (\cref{tab:top50_human_placenta}).
%
Two Ensembl-annotated beta subunits of the chorionic gonadotropin
genes are expressed in \bonobo, one related to \textit{CGB} and
another related to \textit{CGB7}; both are not highly expressed
(\cref{fig:chorionic_gonadotropin}). Among the other top 50 expressed
genes, there is high expression of keratins (e.g., \textit{KRT19},
\textit{KRT18}), heat-shock proteins (e.g., \textit{HSP90B1},
\textit{HSPA1A}), serpins (\textit{SERPINE1}, \textit{SERPINE2}),
insulin growth factor 2 (\textit{IGF2}) and one of its binding
proteins \textit{IGFBP3}, and the hormone adrenomedullin
(\textit{ADM}).
